A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-15 tidyindex.qmd
knit

`Computation`: The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$  as a function of time. In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$,  $\tilde{x}_s^i(t^') = f[[x_s(t)]$ where $t^'$ refers to the new temporal resolution after aggregation. An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$. That is also known as the time scale. 
<!-- Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.  -->
For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).

>>>>

`Computation`: The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$  as a function of time. In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$,  $\tilde{x}_s^i(t^\prime) = f[[x_s(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation. An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$. That is also known as the time scale. 
<!-- Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.  -->
For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).

### 2023-08-15 tidyindex.tex
knit

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners, breakable, interior hidden]}{\end{tcolorbox}}\fi

@@

\begin{itemize}
\item
  \(x^i(s, t)\) represents a variable of interest for example
  precipitation,\(j = 1, \cdots, p\) and
\item
  \(s\) represents the geographic locations in the space
  \(D_s \subseteq \mathbb{R}^m\). Examples of geographic locations
  include a collection of countries, longitude and latitude coordinates
  or regions of interest and,
\item
  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
  For instance, time measurements could be recorded hourly, yearly,
  monthly, quaterly or by season..
\end{itemize}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3514}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}@{}}
\caption{An notation overview of the input, operation, and output of
each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x(t)\) & \(f_{\Psi}[x(.)]\) &
\(x^*(t^\prime)\) where \(t\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x(s)\) & \(g_{\Theta}[x(.)]\) &
\(x^\#(s^\prime)\) where \(s^\prime \in D_{s^\prime}\) \\
3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
???? \\
4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(v(s;t)\)
where bbbbb \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\) where
\(\mathbf{x} \subseteq \mathbb{R}^p\) & aaaaa & \(\mathbf{y}(s;t)\)
where \(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < p\) \\
6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
\(p(.)\) denotes the probability value between 0 and 1 \\
7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
where \(z \sim N(.,.)\) \\
8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
9 & Simplification & \(x(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

\texttt{Computation}: The temporal processing module takes a single
variable in a set of fixed locations, \(x(t)\), as input. This step
transforms the original series into a new series to incorporate the
temporal neighbouring effect: \(x^*(.) = f_{\Psi}[x(.)]\), where
\(\Psi\) is a vector of hyper-parameters associated with the
transformation. In temporal processing, the transition of temporal
dimension from \(t\) to \(t^\prime \subseteq D_{t^\prime}\) allows for
the decomposition of time into its constituent temporal components. For
example, to break down daily data into monthly.

In the Standardized Precipitation Index (SPI), temporal processing is
used to sum the monthly precipitation over a rolling window of size
\(k\), also known as the time scale, prior to the distribution fitting
and normalising module. Assuming an original time series indexed by
\(t= 1, 2, \cdots, T\), the aggregated precipitation series, indexed by
\(t^\prime = 1, 2, \cdots, T-k+1\), can be expressed as
\(x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)\). For SPI,
the choice of the time scale \(k\) is used to control the accumulation
period for the water deficit, enabling the assessment of drought
severity across various types (meteorological, agricultural, and
hydrological).

@@

The transformation can be based on domain-specific knowledge,
originating from theories describing the underlying physical processes,
or guided in statistical methods. The Standardized
Precipitation-Evapotranspiration Index (SPEI) calculates the difference
series (\(D\)) from precipitation (\(P\)) and potential
evapotranspiration (\(\text{PET}\)), using a water balance model
(\(D = P - \text{PET}\)). This is the only step that differentiates it
from the Standardized Precipitation Index (SPI).

When a physical process is absent to describe the phenomenon,
multivariate data are often combined using statistical methods, which
can be further classified into linear and non-linear techniques. Many
social indexes use a linear combination in an additive structure,
\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
is the weight assigned to each individual variable \(x_i\). A linear
combination can also be preceived as a linear projection of multivariate
information through the weight vector. Projecting data from higher to
lower dimension unavoidably leads to information loss and the used
weights need careful examination to understand their effects on the
index and implications for interpretation and decision-making. One can
visualise the data projection across different sets of weights through
animation using a method called tour.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced, frame hidden]}{\end{tcolorbox}}\fi

@@

\begin{itemize}
\item
  \(x^i(s, t)\) represents a variable of interest for example
  precipitation,\(j = 1, \cdots, p\) and
\item
  \(s\) represents the geographic locations in the space
  \(D_s \subseteq \mathbb{R}^m\). Examples of geographic locations
  include a collection of countries, longitude and latitude coordinates
  or regions of interest and,
\item
  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
  For instance, time measurements could be recorded hourly, yearly,
  monthly, quaterly or by season.
\end{itemize}

In what follows when geographical or temporal components of the
\(x(s,t)\) process are fixed we will be using suffix notation. For
example, \(x_{s}(t)\) represents the data for a fixed location \(s\) as
a function of time \(t\). While \(x_{t}(s)\) denotes the spatial varying
procces for a fixed \(t\).

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}@{}}
\caption{An notation overview of the input, operation, and output of
each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x_{s}(t)\) & \(f_{\Psi}[x(.)]\) &
\(\tilde{x}^i(t^\prime)\) where \(t\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x_t(s)\) & \(g[x(.)]\) & \(x^i(s^\prime)\)
where \(s^\prime \in D_{s^\prime}\) \\
3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) &
\(Tx^i_t(s)\) ???? \\
4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(Sx^i(s;t)\)
where bbbbb \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\) where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) & aaaaa &
\(\mathbf{y}(s;t)\) where \(\mathbf{y} \subseteq \mathbb{R}^d\)
and\(d < q\) \\
6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
\(p(.)\) denotes the probability value between 0 and 1 \\
7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
where \(z \sim N(.,.)\) \\
8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
9 & Simplification & \(x(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

\texttt{Computation}: The temporal processing module takes as input
argument a single variable \(x_s^i(t)\) at location \(s\) as a function
of time. In this step the original time series can be transformed or
summarized into a new one via time aggregation, the transformation is
represented by the function \(f\),
\(\tilde{x}_s^i(t^\prime) = f[[x_s(t)]\) where \(t^\prime\) refers to
the new temporal resolution after aggregation. An example of a temporal
processing done in the computation of the Standardized Precipitation
Index (SPI), consists of summing the monthly precipitation series over a
rolling time window of size \(k\). That is also known as the time scale.
For SPI, the choice of the time scale \(k\) is used to control the
accumulation period for the water deficit, enabling the assessment of
drought severity across various types (meteorological, agricultural, and
hydrological).

@@

The transformation can be based on domain-specific knowledge,
originating from theories describing the underlying physical processes,
or guided in statistical methods. For example, the Standardized
Precipitation-Evapotranspiration Index (SPEI) calculates the difference
\(D\) between precipitation (\(P\)) and potential evapotranspiration
(\(\text{PET}\)), using a water balance model (\(D = P - \text{PET}\)).
This is the only step that differs from the Standardized Precipitation
Index (SPI).

(Move this to the example and remove notation) When a physical process
is absent to describe the phenomenon, multivariate data are often
combined using statistical methods, which can be further classified into
linear and non-linear techniques. Many social indexes use a linear
combination in an additive structure,
\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
is the weight assigned to each individual variable \(x_i\). A linear
combination can also be preceived as a linear projection of multivariate
information through the weight vector. Projecting data from higher to
lower dimension unavoidably leads to information loss and the used
weights need careful examination to understand their effects on the
index and implications for interpretation and decision-making. One can
visualise the data projection across different sets of weights through
animation using a method called tour.
